---
title: "Phylogenetic Tree Construction 2"
author: "Mark Watson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "BIOMI4300_Project/figures/PhylogeneticTree/") # send any figure output to this folder 
```

# Goals
1. Load in fastree unrooted tree 
2. Add tree to phyloseq object
3. Visualize Tree
4. Prune ASVs if needed
5. Root Tree
6. Add new tree to phyloseq object
7. Save 2 phyloseq objects 1. unrooted and 2. rooted

# Input
1. fasttree
2. physeq object

# Output
1. rootedphyseq
2. unrooted physeq

# Setting up the Environment
## Set Seed
```{r set-seed}
set.seed(238428)
```

## Timing Script
```{r timing}
start_time <- Sys.time()
start_time

```

## Load Libraries and Phyloseq Object
```{r load-libraries}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools, tidytree,  install = FALSE)

# Load physeq 
load("BIOMI4300_Project/data/processed_physeq.RData")
processed_physeq <- raw_physeq

# load tree
unrooted_tree <- read.tree("BIOMI4300_Project/data/ASVs_unrooted.tree")
unrooted_tree
```

# Merge phyloseq
```{r merge-physeq}
# Intuition check 
stopifnot(ntaxa(processed_physeq) == ntaxa(unrooted_tree))

# Merge the tree with the phyloseq object 
unrooted_physeq <- 
  merge_phyloseq(processed_physeq, unrooted_tree)

```

# plot w/ ggtree
```{r plot-tree}
kingdom_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by kingdom 
  geom_tippoint(mapping = aes(color = Kingdom)) + 
#  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "grey")) +
  # Add title 
  labs(title = "Unrooted Tree") + 
  #move the legend to the bottom 
  theme(legend.position = "bottom"); kingdom_tree

kingdom_node_tree <- 
  kingdom_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)
kingdom_node_tree
```

# Midroot Our Tree
```{r midroot}
# Is the tree rooted?
new_unrooted_tree <- phy_tree(unrooted_tree)

# Check to see if tree is rooted.. expect it to be false
is.rooted(new_unrooted_tree)

# Let's midpoint root the tree
midpoint_rooted_tree <- midpoint.root(new_unrooted_tree)

# Is the new tree rooted? Should be TRUE!
is.rooted(midpoint_rooted_tree)


midpoint_rooted_physeq <- merge_phyloseq(processed_physeq, midpoint_rooted_tree)
#View our final tree

# Plot midrooted simple tree with no text
mid_rooted_simple <- ggtree(midpoint_rooted_physeq) +
  geom_tippoint(aes(color = Phylum))  

mid_rooted_simple

```

# Save phyloseq Object
```{r save-phyloseq}
# Save both phyloseq objects with our tree object to one .RData file 
save(list = c("unrooted_physeq", "midpoint_rooted_physeq"),
     file = "BIOMI4300_Project/data/phytree_preprocessed_physeq.RData")
```

# Session Info
```{r session_info}
devtools::session_info()

```